---
title: "Geog4/6300: Lab 1"
format: gfm
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)

knitr::opts_chunk$set(echo = TRUE)

#Going to be using comments to keep track of what I'm doing! :)
```

## Loading data into R, data transformation, and summary statistics

**Your name: Selena Centeno**

**Overview and lab criteria:**

This lab is intended to assess your ability to use R to load data and to generate basic descriptive statistics. For this lab to be marked complete, the following criteria must be met:

4. Identify and apply appropriate data filtering and cleaning strategies to prepare datasets for analysis. (Task 2)
5. Effectively interpret the code you create, explaining in plain language what each step does to the data. (Task 6)
6. Identify and use appropriate external documentation — including package references, help files, and peer resources — to learn and apply unfamiliar functions or methods. (Task 7)
7. Filter, aggregate, and transform datasets using grouping and summary operations to answer specific analytical questions. (Tasks 1 & 5)
9. Reshape data between wide and long formats to meet the requirements of different analytical or visualization tasks. (Task 4)
11. Create effective data visualizations across multiple chart types (line, scatter, histogram, Q-Q plot), applying appropriate aesthetic choices such as color, grouping, and labeling. (Task 3 & 5)

**Data:**

You'll be using monthly weather data from the Daymet climate database (http://daymet.ornl.gov) for all counties in the United States over an 12-year period (2010-2021). These data are available on the GitHub repo for our course. The following variables are provided:

* `cty_txt`: Code for joining to census data
* `year`: Year of observation (with an initial "Y" to make it a character)
* `month`: Month of observation (1 = Jan, 2 = Feb, etc.)
* `median_tmax`: Median maximum recorded temperature (Celsius)
* `median_tmin`: Median minimum recorded temperature (Celsius)
* `sum_prcp`: Total recorded precipitation for the month (mm)
* `cty_name`: Name of the county
* `state`: State of the county
* `region`: Census region (map: https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf)
* `division`: Census division
* `X`: Longitude of the county centroid
* `Y`: Latitude of the county centroid

These labs are meant to be done collaboratively, but your final submission should demonstrate your own original thought (don't just copy your classmate's work or turn in identical assignments). Your answers to the lab questions should be typed in this Quarto template. You'll then render the document to a GitHub markdown document and upload it to your class GitHub repo.

**Procedure:**

Load the tidyverse package and import the data:

```{r}
#| message: false
#Importing the data:
library(tidyverse)

daymet_data <- read_csv("data/daymet_monthly_median_2010-2021.csv")
```

We can look at the first few rows of the dataset using the *head()* function. We also use *kable* to format the output as a readable table..

```{r}
kable(head(daymet_data))
```

There are a lot of observations here, 452,448 to be exact. To get a better grasp on the data, we can use `group_by()` and `summarise()` from the tidyverse package. This will allow us to identify the mean value for each year by county across the study period.

## Task 1

*Use `group_by()` and `summarise()` to calculate the mean minimum temperature for each year by county across all months, also including State and Region as grouping variables. Your resulting dataset should show the value of tmin for each county in each year. Use the `kable()` and `head()` functions as shown above to call the resulting table.*

```{r}
#| label: task1
tmin_yearly <- daymet_data %>%
  group_by(cty_name, state, region, year) %>%
  summarise(tmin = mean(median_tmin, na.rm = TRUE))
#na.rm tells R to ignore any "NA" values in the dataset. 
kable(head(tmin_yearly))
```

## Task 2

*Let's shift to the state level, focusing on those in the South Region. Filter the original data frame (`daymet_data`) to just include counties in this region. Then calculate the mean minimum temperature by year for each state. For an optional extra challenge, use the `round()` function to include only 1 decimal point. Use `kable()` and `head()` to call the first few lines of the resulting table.*

```{r}
#| label: task2
#Filtering the dataframe to only include countries in the south region using the filter() function. Then, using the groupby function to organize based on states and year. Then, calculating the mean for each group from from the median_tmin column.
southregion_tmin<- daymet_data %>%
  filter(region == "South Region") %>%
  group_by(state, year) %>%
  summarise(tmin=mean(median_tmin, na.rm = TRUE))

#Extra Challenge Part below.
southregion_tminchallenge <- daymet_data %>%
  filter(region == "South Region") %>%
  group_by(state, year) %>%
  summarise(tmin = round(mean(median_tmin, na.rm = TRUE), 1))

kable(head(southregion_tminchallenge))
```

## Task 3

*To visualize the trends, we could use ggplot to visualize change in mean temperature over time. Create a line plot (`geom_line`) showing the state means you calculated in task 2. Use the `color` parameter to show separate colors for each state. You may also need to define the state as a group in the aesthetic parameter.*

```{r}
#| label: task3
#Making pretty plot 
ggplot(southregion_tmin, aes(x = year, y = tmin, color = state, group = state)) +
  geom_line()
```

## Task 4

*If you wanted to look at these data as a table, you'd need to have it in wide format. Use the `pivot_wider()` function to create a wide-format version of the data frame you created in task 2. In this case, the rows should be states, the columns should be the years, and the data in those columns should be mean minimum temperatures. Then call the whole table using `kable()`.*

```{r}
#| label: task4
southregion_tmin_wide <- southregion_tmin %>%
  pivot_wider(names_from = year, values_from = tmin)

kable(southregion_tmin_wide)
#Looks nice. Rows are the states and columns are years. I'm going to make the mean tmin temps rounded like I did in Task 2!

southregion_tminchallengewide <- southregion_tmin %>%
  mutate(tmin = round(tmin, 1)) %>%
  pivot_wider(names_from = year, values_from = tmin)

kable(southregion_tminchallengewide)
```

## Task 5

*Let's assess the relationship of heat and precipitation by region. Returning to the original dataset, create a data frame that shows the mean maximum temperature and mean precipitation for all states in 2015, also including region as a subgroup in your `group_by`. Then use ggplot to create a scatterplot (`geom_point`) for these two variables, coloring the points using the region variable.*

```{r}
#| label: task5
#First making the data frame for the mean max temp and mean precip, filtered to only 2015. Then grouping like we did earlier in the lab.
mmtemp_mprcp_2015 <- daymet_data %>%
  filter(year == "Y2015") %>%
  group_by(state, region) %>%
  summarise(mean_tmax = mean(median_tmax, na.rm = TRUE), mean_prcp = mean(sum_prcp, na.rm = TRUE))
#For some reason the "mean_tmax" column is all "NA." I'm going to try the as.numeric function to see if it'll fix it.
daymet_data <- daymet_data %>%
  mutate(median_tmax = as.numeric(median_tmax))
#YES IT DOES!
#------------------------------------------------
#Now make the pretty ggplot.
ggplot(mmtemp_mprcp_2015, aes(x = mean_tmax, y = mean_prcp, color = region)) +
  geom_point()
```

## Task 6

*In the space below, explain what each function in your code for task 5 does to the dataset in plain English.*

The "filter()" function cleans the dataset down to only show observations from the year 2015 (or Y2015 in the dataframe). The "group_by()" function organizes the observations from 2015 (Y2015) by state and region so that the calculations in future functions can be done separately for each state. The "summarise()" function is condensing the 2015 observations for each state into a single row that contains the averages for median_tmax. Within the "summarise()" function is "mean(), which calculates the average tmax and average precipitation. The "na.rm = TRUE" tells R to ignore any missing values when its doing the calculations. When making the plot, "ggplot()" is the base function to which what graph is going to be made from the data (a scatterplot in this case). The "aes()" function assigns mean tmax to the x axis and mean precipitation to the y-axis. The "color" of each point is determined by the region. the "geom_point()" function tells R to display the states as points on the scatterplot.

## Task 7

*The `dplyr` package also includes `across` function. Use `?across` on the R command line to open the documentation for this function. In the space below, explain what it does in your own words. Then interpret the way the across function is used below, going line by line within the function.*

```{r}
state_2015 <- daymet_data %>%
  filter(year == "Y2015") %>%
  group_by(region, state) %>%
  summarise(
    across(
      c(median_tmax, sum_prcp),
      mean,
      na.rm = TRUE,
      .names = "mean_{.col}"
    )
  )
?across
```


The "across()" functions makes it so you can apply the same function to multiple rows at the same time. Rather than writing out a calculation separately for each column. you can use across to select multiple to apply the same function. In the code block above, within the across() function, "c(median_tmax, sun_prcp) selects those two columns that the function is going to be applied to. The "mean" part is the calculation for each column selected. na.rm = TRUE tells R to ignore any missing values when performing the calculation. .names = "mean_{.col} tells R how to name the new calculated columns.


## Challenge Question

In class, we covered ways of working with the Daymet API. Create a script below that uses the **daymetr** package to download data from Daymet for a place (or places) of your choosing. Then visualize the temporal pattern for a variable of your choosing in this place, similar to what you did in question 4. Use a dplyr function (`mutate()`, `summarise()`, `filter()`, etc.) to do any needed data wrangling and create a visual using ggplot.

In addition to this code, write a short summary of a pattern that's evident in the data you visualized.

```{r}
#| label: challenge

# Your code goes here
```

*Explanation goes here.*

## Final Submission Stuff

### Disclosure of Assistance

I used AI assistance (ChatGPT) to tell me what exact R is doing when using certain functions when I couldn’t remember what the core meaning was for a function. Using the AI assistance supported my learning of this new coding language as I am now able to understand what R is doing behind the scenes when I input certain functions. It is a great search engine when needing a quick answer. 

### Lab Reflection
On a scale of 1-10, I'd say a 4, and the work on this lab was between easy and moderate. I had some trouble at the beginning of the lab with a few errors that I had to work around, but I eventually got them figured out. Had to slow down and really learn/soak in what certain functions are doing like the last lab. 